\documentclass[aip,jcp,amsmath,amssymb,preprint]{revtex4-2}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[version=4]{mhchem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,fit,positioning}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}

\begin{document}

\author{Yuan Jiao}
\email[]{jiaoyuan24@mails.ucas.ac.cn}
\affiliation{School of Advanced Interdisciplinary Sciences,
 University of Chinese Academy of Sciences, Beijing, China}

\title{GradTDDFT: A Differentiable Time-Dependent Density Functional
 Theory Framework for Training Neural Network
 Exchange-Correlation Functionals}

\date{\today}

\begin{abstract}
GradTDDFT is a JAX-based framework for differentiable
time-dependent density functional theory (TDDFT) with neural
exchange-correlation (XC) functionals. The method separates a
fixed molecular data layer---basis functions, quadrature grids,
one-electron integrals, and electron-repulsion integrals---from
the differentiable Kohn--Sham SCF and TDDFT response layers. This
separation gives a well-defined computational graph from neural
XC parameters to self-consistent orbitals, TDA/Casida response
matrices, excitation energies, and oscillator strengths. The central technical element
is an energy-consistent local response tensor: the XC potential and
$f_{\rm xc}$ kernel are obtained as first and second derivatives
of the same learned local energy density by JAX automatic
differentiation. The neural functional uses a coefficient-basis
architecture in which pointwise density, gradient,
kinetic-energy-density, local exact-exchange, and optional MP2/PT2
descriptors produce mixing fields over fixed semilocal, HF, and
MP2/PT2 channels. We formulate the resulting ground-state and
excited-state objectives together with
the corresponding computational evaluation protocol.
\end{abstract}

\keywords{differentiable programming, JAX, TDDFT,
neural exchange-correlation functional, excited states}

\maketitle

\input{sections/introduction}
\input{sections/theory}
\input{sections/implementation}
\input{sections/results}
\input{sections/discussion}
\input{sections/conclusion}

\begin{acknowledgments}
The authors thank the developers of PySCF, JAX, Flax, Optax, and
GradDFT for providing the foundational tools upon which GradTDDFT
is built.
\end{acknowledgments}

\section*{Supplementary Material}

The supplementary material contains the following files.
\begin{itemize}
  \item Detailed theory and response-kernel derivations
  \item Computational input records for the molecular benchmarks
  \item Reference geometries, grids, basis choices, and convergence
    settings used in the reported calculations
  \item High-resolution versions of the figures in the main text
\end{itemize}

\section*{Code Availability}

GradTDDFT is available under the Apache 2.0 license on
\href{https://github.com/STOKES-DOT/TD-GradDFT}{GitHub}.
The released code contains the research implementation and supporting
materials for this manuscript. Numerical values should be interpreted
together with the recorded geometry, basis, grid, convergence, backend,
and reference settings.

\bibliography{references}

\end{document}
