\section{Introduction}

Time-dependent density functional theory (TDDFT) within the
linear-response formalism of Casida\cite{Casida1995} is one of
the standard approaches for computing molecular excitation
energies and oscillator strengths. Its cost is often compatible
with chemical screening studies, but its accuracy is controlled
by the ground-state exchange-correlation (XC) functional and by
the associated response kernel. Conventional semilocal and hybrid
functionals therefore inherit known limitations for
charge-transfer excitations, Rydberg states, static correlation,
and double-excitation character. These failures are not merely
post-processing errors: the same density functional determines
the SCF density, orbital energy gaps, the XC potential, and the
kernel entering the TDDFT response matrices.

Neural network XC functionals offer a route to learn more flexible
functional forms, including nonlocal XC structure, from accurate
reference data.\cite{GLOBE,Schmidt2019NonlocalXCML}
For excited states, however, fitting ground-state energies alone is
not enough: a functional used in TDDFT must also produce stable
self-consistent densities, useful frontier orbital gaps, and a
response kernel consistent with the learned XC energy. A recent
overview of differentiable DFT emphasizes that this changes the
role of the KS equations: the functional is not only evaluated inside
a self-consistent DFT calculation, but trained through the
differentiable SCF map, with gradients from observables propagated
back to the model parameters while the governing electronic-structure
equations remain in the loop.\cite{vonStrachwitz2026DifferentiableDFT}
For TDDFT, the corresponding differentiable program must extend
beyond the ground-state SCF fixed point to include response-kernel
construction, TDA/Casida eigenvalue problems, excitation energies,
oscillator strengths, and spectrum-level objectives. Differentiable
training is therefore important because it can place losses on the
quantities actually used to judge excited-state predictions, rather
than only on pre-response ground-state energies.
This differentiable view also changes how the XC
derivatives are implemented. In conventional DFT codes, the XC
potential $v_{\rm xc}=\delta E_{\rm xc}/\delta\rho$ and the
TDDFT kernel $f_{\rm xc}=\delta^2E_{\rm xc}/\delta\rho\,\delta\rho'$
are usually derived analytically for each functional form and then
encoded as separate routines. GradTDDFT instead treats the local XC
energy density as the primary differentiable object and obtains these
first- and second-order derivatives by automatic differentiation,
so the SCF potential and the response kernel remain tied to the same
energy expression.\cite{Ekstrom2010ADResponse}

This idea is part of a broader movement toward differentiable
scientific codes. At the level of methods and software, D4FT and
DQC expose differentiable DFT workflows,\cite{D4FT,DQC} PySCFAD
differentiates PySCF-style mean-field and post-mean-field
calculations,\cite{PySCFAD} GradDFT focuses on machine-learning
enhanced ground-state density functionals,\cite{GradDFT} and jax-xc
translates conventional Libxc functional forms into JAX-compatible
energy-density routines.\cite{JAXXC} Differentiable semiempirical
models provide a complementary low-cost route: dxtb implements
differentiable extended tight-binding, and related semiempirical
formulations have reported ground- and excited-state
gradients.\cite{DXTB,Athavale2026DifferentiableSemiempirical}
At the level of molecular and materials applications, JAX-MD provides
differentiable simulation primitives for molecular systems,\cite{JAXMD}
JAX-based DFT workflows have been developed for crystalline systems
and direct solid-state optimization,\cite{JRystal,JRystalDirectOptimization}
plane-wave AD-DFPT has been applied to materials design, error
control, and parameter learning,\cite{Schmitz2025ADDFPT} and
orbital-free DFT has used automatic differentiation to obtain
functional potentials and higher-order material properties directly
from energy expressions.\cite{Tan2023OFDFTAD} These developments
demonstrate that automatic differentiation can turn molecular
simulation and electronic-structure calculations into trainable
numerical programs.
The missing piece for excited states is an end-to-end differentiable
TDDFT formulation in which the neural XC energy, SCF potential,
response kernel, and excited-state observables belong to the same
differentiable computational graph.

We present GradTDDFT to address this gap. The framework keeps PySCF
as the source of molecular construction, basis information, grids, and
integrals, while expressing the SCF, XC, TDDFT response, and training
logic in JAX. GradTDDFT separates fixed molecular data from the
differentiable KS-SCF and linear-response layers, giving a continuous
map from neural XC parameters to self-consistent densities, orbital
rotations, TDA/Casida matrices, excitation energies, oscillator
strengths, and excited-state objectives. The response eigenproblem is
solved by a Davidson iteration in the forward pass and differentiated
using the implicit eigensolver strategy of Lei Wang and
co-workers,\cite{Xie2020DominantEigensolverAD} so gradients are reconstructed
from the converged Ritz pairs rather than from the full iterative solver
history. The neural
functional is written as a coefficient-basis family in which local
density descriptors generate mixing fields over semilocal, HF, and
optional MP2/PT2 channels. For semilocal channels, the XC energy,
potential, and local $f_{\rm xc}$ tensor are obtained from the same
energy-density expression. The HF channel supplies a scalar hybrid
exchange response, while the MP2/PT2 channel supplies a
root-specific second-order correction after the TDA/Casida singles
problem is solved. This design provides separate ground-state and
excited-state training protocols within the same differentiable
implementation.
