\section{Theory and Methods}
\label{sec:theory}

% ======================================================================
\subsection{Differentiable Kohn--Sham SCF}
\label{sec:scf}

The Kohn--Sham (KS) equations for a restricted closed-shell system
are the single-particle eigenvalue equations
\begin{equation}
  \left[-\frac{1}{2}\nabla^2 + v_{\rm ext}(\mathbf{r})
  + v_J(\mathbf{r}) + v_{\rm xc}(\mathbf{r})\right]
  \phi_i(\mathbf{r}) = \epsilon_i \phi_i(\mathbf{r}),
  \label{eq:ks}
\end{equation}
where $v_{\rm xc}(\mathbf{r}) = \delta E_{\rm xc} /
\delta\rho(\mathbf{r})$ is the XC potential and
$\rho(\mathbf{r}) = 2\sum_i |\phi_i(\mathbf{r})|^2$.

In the atomic orbital (AO) basis $\{\chi_\mu\}$, these become
$\mathbf{F}\mathbf{C} = \mathbf{S}\mathbf{C}\boldsymbol{\epsilon}$,
with the Fock matrix
\begin{equation}
  \mathbf{F} = \mathbf{h} + \mathbf{J}[\mathbf{D}]
  + \mathbf{V}_{\rm xc}[\mathbf{D}],
  \label{eq:fock}
\end{equation}
solved iteratively via the SCF cycle
\begin{equation}
  \mathbf{D}^{(n)} \rightarrow \mathbf{F}^{(n)}
  \rightarrow \mathbf{C}^{(n+1)} \rightarrow \mathbf{D}^{(n+1)},
  \label{eq:scf-cycle}
\end{equation}
with DIIS acceleration.\cite{DIIS}
GradTDDFT implements both restricted (RKS) and unrestricted (UKS)
solvers.

\emph{Differentiable SCF modes.} GradTDDFT provides two modes for
gradient propagation through the SCF cycle:

\begin{itemize}
  \item \textbf{Fixed-density mode.} A converged reference density is
    held fixed via \texttt{jax.lax.stop\_gradient}; only XC
    functional parameters $\theta$ enter the computational graph.
    This is computationally efficient and numerically stable for
    early-stage training where the density is expected to remain close
    to the reference.

  \item \textbf{Self-consistent mode.} The full SCF cycle enters the
    computational graph. Gradients propagate through the implicit
    fixed-point iteration via an implicit commutator approach---where
    the SCF residual $[\mathbf{F}, \mathbf{D}]$ is driven to zero and
    gradients are obtained through the implicit function theorem---or
    via explicit unrolled differentiation. The implicit commutator
    mode is the recommended default due to superior robustness across
    diverse molecular configurations.
\end{itemize}

% ======================================================================
\subsection{Linear-Response TDDFT}
\label{sec:tddft}

Within the linear-response formalism, excitation energies
$\Omega_I$ and transition amplitudes are obtained by solving the
Casida eigenvalue equation:\cite{Casida1995}
\begin{equation}
  \begin{pmatrix}
  \mathbf{A} & \mathbf{B} \\
  \mathbf{B}^* & \mathbf{A}^*
  \end{pmatrix}
  \begin{pmatrix} \mathbf{X}_I \\ \mathbf{Y}_I \end{pmatrix}
  = \Omega_I
  \begin{pmatrix}
  \mathbf{1} & \mathbf{0} \\
  \mathbf{0} & -\mathbf{1}
  \end{pmatrix}
  \begin{pmatrix} \mathbf{X}_I \\ \mathbf{Y}_I \end{pmatrix},
  \label{eq:casida}
\end{equation}
where $\mathbf{X}_I$ and $\mathbf{Y}_I$ are the excitation and
de-excitation amplitudes in the occupied-virtual MO rotation basis.

For restricted closed-shell systems, the orbital Hessian matrix
elements are
\begin{equation}
  \begin{aligned}
  A_{ia,jb} &=
    (\epsilon_a - \epsilon_i)\delta_{ij}\delta_{ab}
    + 2(ia|jb) - c_x(ij|ab) + f_{ia,jb}^{\rm xc}, \\
  B_{ia,jb} &=
    2(ia|bj) - c_x(ib|aj) + f_{ia,jb}^{\rm xc,B},
  \end{aligned}
  \label{eq:ab-matrix}
\end{equation}
where $(pq|rs)$ are two-electron integrals in the MO basis,
$c_x$ is the Hartree--Fock exchange fraction, and
$f_{ia,jb}^{\rm xc}$ is the exchange-correlation kernel
contribution (see Section~\ref{sec:fxc}). Indices $i,j$ denote
occupied MOs, $a,b$ virtual MOs.

The Tamm--Dancoff approximation (TDA) sets $\mathbf{B}=0$,
reducing Eq.~\ref{eq:casida} to the Hermitian eigenvalue problem
$\mathbf{A}\mathbf{X}_I = \Omega_I \mathbf{X}_I$.

From the solution vectors, physical observables are computed:
\begin{align}
  \boldsymbol{\mu}_I &=
    \sum_{ia} (X_{ia,I} + Y_{ia,I})
    \boldsymbol{\mu}_{ia}, \label{eq:transdip} \\
  f_I &= \frac{2}{3}\Omega_I |\boldsymbol{\mu}_I|^2,
    \label{eq:oscstr}
\end{align}
where $\boldsymbol{\mu}_{ia} = \langle i|\mathbf{r}|a\rangle$ are
MO transition dipole integrals, and $f_I$ are oscillator strengths.

% ======================================================================
\subsection{Strict $f_{\rm xc}$ Kernel via JAX Automatic Differentiation}
\label{sec:fxc}

The exchange-correlation kernel is the second functional derivative
of the XC energy with respect to the density:
\begin{equation}
  f_{\rm xc}(\mathbf{r}, \mathbf{r}') =
    \frac{\delta^2 E_{\rm xc}}
         {\delta\rho(\mathbf{r})\delta\rho(\mathbf{r}')}.
  \label{eq:fxcfd}
\end{equation}

In most quantum chemistry packages, $f_{\rm xc}$ is implemented
through analytic derivatives of approximate XC functional forms
or via finite differences. For neural XC functionals, where the
functional form is a neural network with thousands of parameters,
analytic derivatives are infeasible and finite differences introduce
numerical noise that compromises gradient-based optimization.

GradTDDFT addresses this through the \emph{strict local response
tensor} approach. The XC energy is expressed as an integral over
a local energy density:
\begin{equation}
  E_{\rm xc} = \int d\mathbf{r}\,
    e_{\rm xc}^{\rm loc}\bigl(u(\mathbf{r})\bigr),
  \label{eq:eloc}
\end{equation}
where $u(\mathbf{r})$ represents the set of reduced local variables
at grid point $\mathbf{r}$. For generalized gradient approximation
(GGA)-type functionals,
$u = [\rho, \partial_x\rho, \partial_y\rho, \partial_z\rho]$.

The $f_{\rm xc}$ contribution to the response matrices is then
constructed by contracting a grid-based local Hessian with
transition response features:
\begin{equation}
  K_{ia,jb}^{\rm xc} = 2\sum_g w_g
    \sum_{\mu\nu}
    H_{\mu\nu}(\mathbf{r}_g)
    B_{ia}^\mu(\mathbf{r}_g)
    B_{jb}^\nu(\mathbf{r}_g),
  \label{eq:fxccontraction}
\end{equation}
where
\begin{align}
  H_{\mu\nu}(\mathbf{r}) &=
    \frac{\partial^2 e_{\rm xc}^{\rm loc}}
         {\partial u_\mu(\mathbf{r})\, \partial u_\nu(\mathbf{r})},
    \label{eq:localhessian} \\
  B_{ia}^\mu(\mathbf{r}) &=
    \frac{\partial u_\mu(\mathbf{r})}{\partial \kappa_{ia}},
    \label{eq:transresponse}
\end{align}
and $\kappa_{ia}$ are occupied-virtual orbital rotation parameters.

The core innovation is that GradTDDFT computes the local Hessian
$H_{\mu\nu}(\mathbf{r})$ entirely via JAX automatic
differentiation:
\begin{quote}
\texttt{jax.hessian(}\\
\quad\texttt{self.\_total\_point\_local\_energy\_from\_variables,}\\
\quad\texttt{argnums=1)}\\
\texttt{)}
\end{quote}
This eliminates finite-difference errors, correctly handles
the cross-derivatives between semilocal coefficients and density
variables, and remains fully differentiable with respect to
neural XC parameters $\theta$, enabling gradient flow through
the entire TDDFT pipeline during training.

For hybrid functionals with HF exchange, the exact exchange kernel
is treated separately as a strict nonlocal contribution:
\begin{equation}
  K_{ia,jb}^{\rm HF,A} = -c_x(ij|ab), \qquad
  K_{ia,jb}^{\rm HF,B} = -c_x(ib|aj),
  \label{eq:hfkernel}
\end{equation}
where $c_x$ may be a global constant or a density-weighted effective
exchange fraction for neural hybrid functionals.

% ======================================================================
\subsection{Neural XC Functional Architecture}
\label{sec:neuralxc}

The neural XC functional in GradTDDFT follows the
\emph{coefficient-basis} paradigm, in which a pointwise neural
network maps local density descriptors to mixing coefficients that
contract a fixed basis of semilocal XC channels. The local XC
energy density at grid point $\mathbf{r}_g$ is

\begin{equation}
  e_{\rm xc}^{\rm loc}(\mathbf{r}_g) =
    \sum_{k=1}^{n_{\rm sl}}
    c_k^\theta(u_g)\, e_k^{\rm sl}(u_g)
    + f_{\rm pt2}^\theta(u_g)\, \bar{p}_g
    + f_x^\theta(u_g)\, \bar{h}_g,
  \label{eq:neural-eloc}
\end{equation}

and the total XC energy is obtained by numerical quadrature:
\begin{equation}
  E_{\rm xc}^\theta = \sum_g w_g\, e_{\rm xc}^{\rm loc}(\mathbf{r}_g).
  \label{eq:totalexc}
\end{equation}

\textbf{Semilocal basis channels.} The fixed basis
$\{e_k^{\rm sl}\}$ supplies the physically-motivated density
functionals that the neural network reweights. In the default
B3LYP-style configuration these are
$\{e_x^{\rm LDA}, e_x^{\rm B88}, e_c^{\rm VWN-RPA},
 e_c^{\rm LYP}\}$, providing separate exchange and correlation
channels for the network to interpolate.

\textbf{Input descriptors.} The local variable vector $u_g$
follows the DM21-style feature set:\cite{DM21} the electron density
$\rho$, its gradient components and norm $|\nabla\rho|$, and
optionally the kinetic energy density $\tau$. For GGA-like
functionals the reduced variable set is
$u = [\rho, \partial_x\rho, \partial_y\rho, \partial_z\rho]$. All
descriptors are evaluated pointwise on a numerical integration grid
imported from PySCF.

\textbf{Pointwise mixing network.} The coefficient fields
$\{c_k(u)\}$, $f_{\rm pt2}(u)$, and $f_x(u)$ are produced by a
shared MLP that processes each grid point independently. GradTDDFT
provides two MLP architectures:

\begin{itemize}
  \item \textbf{Residual Mixing MLP.} Hidden layers are augmented
    with residual (skip) connections, which improve gradient flow
    through deeper networks. Each residual block applies
    Dense $\rightarrow$ GELU activation $\rightarrow$ Dense, with
    the input added to the output before the nonlinearity.

  \item \textbf{Simple Mixing MLP.} A standard feed-forward MLP
    without residual connections. Lighter weight, suitable for
    rapid prototyping and small-basis configurations.
\end{itemize}

Both architectures use the GELU activation function throughout and
are implemented in JAX/Flax.\cite{Flax} The output layer produces
$n_{\rm sl} + 2$ scalar values per grid point: one coefficient
per semilocal channel, plus the HF and PT2 mixing fields.

\textbf{HF and PT2 mixing heads.} The auxiliary fields
$\bar{h}_g$ and $\bar{p}_g$ in Eq.~\ref{eq:neural-eloc} are
frozen local surrogates for the exact-exchange and second-order
perturbation theory (PT2) energy densities, respectively. The
learned mixing fields $f_x^\theta(u)$ and $f_{\rm pt2}^\theta(u)$
act as smooth, unit-interval weighting functions that control the
local admixture of these nonlocal correlation channels. For hybrid
functionals, the HF mixing field is density-averaged to produce an
effective global exchange fraction:
\begin{equation}
  \alpha_{\rm eff}^\theta =
    \frac{\sum_g w_g\, \rho_g\, f_x^\theta(u_g)}
         {\sum_g w_g\, \rho_g},
  \label{eq:alpha-eff}
\end{equation}
which enters the nonlocal HF kernel
(Eq.~\ref{eq:hfkernel}).

\textbf{Differentiable XC potential.} The XC potential
$v_{\rm xc}(\mathbf{r})$ required by the Fock matrix
(Eq.~\ref{eq:fock}) is obtained as the first derivative of the
local energy density with respect to the density variables:
\begin{equation}
  v_{\rm xc}^\mu(\mathbf{r}_g) =
    \frac{\partial e_{\rm xc}^{\rm loc}}
         {\partial u_\mu(\mathbf{r}_g)}.
  \label{eq:vxc}
\end{equation}
In GradTDDFT this derivative is computed automatically via JAX
reverse-mode differentiation (\texttt{jax.grad}), ensuring that
the XC potential is always exactly consistent with the energy---a
key advantage over implementations that require manually derived
potential formulas. The second derivative (the $f_{\rm xc}$ kernel,
Eq.~\ref{eq:localhessian}) is obtained by applying
\texttt{jax.hessian} to the same local energy function, as
described in Section~\ref{sec:fxc}.

% ======================================================================
\subsection{Training Paradigm}
\label{sec:training}

GradTDDFT adopts a two-stage training strategy that leverages
the full differentiability of the SCF-to-observables pipeline.

\textbf{Stage 1: Ground-state pre-training.} The neural XC
functional is first trained against ground-state reference data,
typically the total energy and/or the real-space electron density:
\begin{equation}
  \mathcal{L}_{\rm GS} =
    w_E \cdot \text{MSE}(E^\theta, E^{\rm ref})
    + w_\rho \cdot \text{MAE}(\rho^\theta, \rho^{\rm ref}),
  \label{eq:gsloss}
\end{equation}
where $E^\theta$ is the SCF total energy obtained with the neural
functional and $\rho^\theta$ is the corresponding self-consistent
density. This stage uses the self-consistent SCF mode with the
implicit commutator gradient, which captures the density response
to functional changes. For rapid prototyping, the fixed-density mode
can be substituted; it uses a frozen reference density for the XC
evaluation and propagates gradients through the XC energy only.

\textbf{Stage 2: Excited-state fine-tuning.} After ground-state
pre-training, the functional parameters are further optimized
directly against excited-state properties:
\begin{equation}
  \mathcal{L}_{\rm ES} =
    w_\Omega \cdot \text{MAE}(\Omega_I^\theta, \Omega_I^{\rm ref})
    + w_f \cdot \text{MAE}(f_I^\theta, f_I^{\rm ref}),
  \label{eq:esloss}
\end{equation}
where $\Omega_I$ and $f_I$ are the excitation energies and
oscillator strengths for selected excited states $I$. The
gradient pathway during fine-tuning is end-to-end: changes in
neural XC parameters $\theta$ modify the local energy density
$e_{\rm xc}^{\rm loc}$ (Eq.~\ref{eq:neural-eloc}), the XC
potential $v_{\rm xc}$ (Eq.~\ref{eq:vxc}), the self-consistent
density $\rho$, the orbital Hessian matrices $\mathbf{A}$ and
$\mathbf{B}$ (Eq.~\ref{eq:ab-matrix}) through the $f_{\rm xc}$
kernel (Eq.~\ref{eq:fxccontraction}), the Casida eigensolution
(Eq.~\ref{eq:casida}), and ultimately the predicted excitation
energies and oscillator strengths. All gradients are computed
automatically by JAX without manual derivation of any intermediate
derivatives.
